Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2019-03-28 article.tex

Although the problem of parameter estimation in non-linear ODE models is highly relevant and at the heart of statistical computing, there are currently not more than four \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN), namely \pkg{FME} \citep{FME}, \pkg{nlmeODE} \citep{nlmeODE}, \pkg{mkin} \citep{mkin} and \pkg{scaRabee} \citep{scaRabee}. All packages have in common that they are built upon the \pkg{deSolve} package \citep{deSolve}. Taking a broader perspective on the topic of dynamic modeling and inference, we find more packages, dealing with, e.g., discrete-time and continuous-time stochastic systems \citep{CollocInfer} or the statistical inference of partially observed Markov processes \citep{pomp}.

@@

\subsection{Simulation and prediction}
These processes involved in bile acid transport give rise to the differential equations and corresponding flowchart presented in Figure~\ref{fig:flowchart}. 
\begin{figure}[ht]
	\centering
	\begin{minipage}{0.55\textwidth}
		\textbf{A}
		\begin{equation*}
			\begin{aligned}
				\frac{\texttt d}{ \texttt{dt} } \verb!TCA_buffer! &\texttt{ = } \verb!-import * TCA_buffer!\\
				& \quad \verb! + export_sinus * TCA_cell!\\
				& \quad \verb! + reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cana! &\texttt{ = } \verb!export_cana * TCA_cell!\\
				& \quad \verb! - reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cell! &\texttt{ = } \verb!import * TCA_buffer!\\
				& \quad \verb! - export_sinus * TCA_cell!\\
				&\quad \verb! - export_cana * TCA_cell!.
			\end{aligned}
		\end{equation*}
	\end{minipage}
	\begin{minipage}{0.44\textwidth}
		\textbf{B}\\
	\includegraphics[width = \textwidth]{images/flowchart}
	\end{minipage}
	\caption{Differential equations and flowchart of the reaction network. Taurocholic acid,  TCA, is transported between three compartments by four different processes. (A) Assuming mass-action kinetics, the three dynamic states satisfy a set of coupled differential equations. (B) The equations are visualized in a flowchart. }
	\label{fig:flowchart}
\end{figure}
Each transportation process is modeled by mass-action kinetics. A possible implementation in \pkg{dMod} is:
\begin{CodeChunk}
	\begin{CodeInput}
# Add reactions
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell",
			 rate = "import*TCA_buffer",
			 description = "Uptake")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer",
			 rate = "export_sinus*TCA_cell",
			 description = "Sinusoidal export")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana",
			 rate = "export_cana*TCA_cell",
			 description = "Canalicular export")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer",
			 rate = "reflux*TCA_cana",
			 description = "Reflux into the buffer")

# Translate into ODE model
mymodel <- odemodel(reactions, modelname = "bamodel")

# Generate prediction function from ODE model
x <- Xs(mymodel, condition = NULL)
	\end{CodeInput}
\end{CodeChunk}
The reactions are collected in an \code{eqnlist} object. The \code{odemodel()} command composes the single reactions to an ODE system and auto-generates the \proglang{C} code which is used by the \pkg{deSolve} package to evaluate the ODE. Prediction functions are generated by the \code{Xs()} command. The usage of the prediction function is illustrated by the following code chunk. Time points are defined between 0 and 50, numeric values are assigned to all model parameters. Finally the prediction function is called and both, the prediction and the sensitivities, are plotted, shown in Figure~\ref{fig:prediction}.

\begin{CodeChunk}
\begin{CodeInput}
times <- seq(0, 50, .1)
pars <- c(TCA_buffer = 1,          # initial values
          TCA_cell = 0,            # initial values
	  TCA_cana = 0,            # initial values
	  import = 0.2,
	  export_sinus = 0.2,
	  export_cana = 0.04,
	  reflux = 0.1)

# Generate the prediction
out <- x(times, pars)
plot(out)

# Get sensitivities for the rate parameters only, pars[4:7]
out <- getDerivs(x(times, pars[4:7], fixed = pars[1:3]))
plot(out)
\end{CodeInput}
\end{CodeChunk}

@@

First, observables are defined by an \code{eqnvec} object from which an observation function \code{g} is generated by the \code{Y()} command. The \code{Y()} command needs to be informed which of the symbols are variables (dynamic states) or parameters. Conveniently, \code{Y()} can retrieve this information from the prediction function \code{x} it is built on, or parse an \code{eqnvec} or \code{eqnlist} such as \code{reactions}.
%For convenience, this can be done passing the \code{reactions} that underly our ODE model.
\begin{CodeChunk}
\begin{CodeInput}
# Generate observation function
observables <- eqnvec(
  buffer = "s*TCA_buffer",
  cellular = "s*(TCA_cana + TCA_cell)"
)

g <- Y(observables, f = x, condition = NULL,
       compile = TRUE, modelname = "obsfn")

\end{CodeInput}
\end{CodeChunk}
Observation functions link internal to observable states. Thus, providing values for the model parameters, the observation function can be used to simulate the outcome of an experiment. Adding noise to the prediction, experimental data is simulated.
In the following, we will simulate the outcome of an \textit{efflux experiment}.
The experiment starts with all TCA concentrations in steady state, such as shown in Figure~\ref{fig:prediction} after $t = 50$. To initiate the efflux, the buffer is replaced by TCA-free buffer, i.e.~\code{TCA_buffer = 0}. This translates into the following initial parameter values:
%, i.e. time point $t = 0$ reflects the distribution of TCA after uptake, as shown in Figure~\ref{fig:prediction} after $t = 50$. Hence, we reset some of the parameter values. Initial values are set to steady state values. The incubation buffer is replaced by fresh buffer, meaning that \code{TCA_buffer} is set to 0. Finally, we assume that \code{s = 1000}. The prediction of the observables is made by concatenation of \code{g} and \code{x} via the concatenation operator \code{"*"} being defined in \pkg{dMod} for prediction, observation and parameter functions.
\begin{CodeChunk}
\begin{CodeInput}
# Reset parameter values
pars["TCA_cell"] <- 0.3846154
pars["TCA_cana"] <- 0.1538462
pars["TCA_buffer"] <- 0
pars["s"] <- 1e3
\end{CodeInput}
\end{CodeChunk}
The predicted dynamics of the system's internal and observable states is obtained by evaluation of the concatenated prediction function $g\circ x$, formulated as \code{g*x} in \pkg{dMod}. The scaling parameter \code{s} is set to 1000.
\begin{CodeChunk}
\begin{CodeInput}
out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Since \code{g} and \code{x} have been generated as generic functions, i.e.~\code{condition = NULL}, we can assign the output to a condition of our choice, in this case ``standard''.
The predicted noiseless observation is obtained by considering the observable states only at the time point of observation \code{timesD}.
%Finally, we obtain the simulated data by taking a subset of the prediction.
\begin{CodeChunk}
\begin{CodeInput}
# Simuate data
timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
datasheet <- subset(as.data.frame(out),
                    time %in% timesD & name %in% names(observables))
\end{CodeInput}
\end{CodeChunk}

>>>>

Although the problem of parameter estimation in non-linear dynamic systems is highly relevant and at the heart of statistical computing, to this day there are not more than four \proglang{R} packages published on the topic on the Comprehensive \proglang{R} Archive Network (CRAN), namely \pkg{FME} \citep{FME}, \pkg{nlmeODE} \citep{nlmeODE}, \pkg{mkin} \citep{mkin} and \pkg{scaRabee} \citep{scaRabee}. All packages have in common that they are built upon the \pkg{deSolve} package \citep{deSolve}.
The packages \pkg{mkin} and \pkg{FME} support ODEs defined by compiled code where \pkg{mkin} also provides tools to autogenerate the \proglang{C} code and compile it using the \pkg{inline} package. For model fitting and uncertainty analysis, \pkg{mkin} fully resorts to the functionality of \pkg{FME}.
Concerning model fitting, \pkg{FME} and \pkg{nlmeODE} (with \pkg{nlme} in the background) use deterministic derivative-based optimizers by default, i.e.~either Levenberg-Marquardt or Newton methods provided by \code{nls.lm()}, \code{optim()} or \code{nlmimb()}. Although all these optimizers support gradient or Hessian information as input, only the \pkg{nlmeODE} package provides an option to augment the ODE by its sensitivity equations to generate derivates for the residuals. By default, sensitivities are computed by finite differences. Last but not least, the \pkg{scaRabee} package uses the Nelder-Mead optimization algorithm which is derivative-free but needs in general more iterations until convergence compared to derivative-based methods.
In all packages, uncertainty analysis is by default based on the variance-covariance matrix, i.e.~on the inverse Hessian matrix of the least squares function. For non-linear models, this method provides a good approximation only if parameters are identifiable and the data is highly informative. If these conditions are not met, more sophisticated methods like e.g.~Markov chain Monte Carlo (MCMC) sampling, implemented in \pkg{FME}, are required.
The strength of \pkg{scaRabee} and especially \pkg{nlmeODE} is multi-conditional fitting. This means that the same model with the same parameters but different forcings to reflect experimental conditions is fitted simultaneously to the condition-specific data sets. In the context of mixed-effects modeling as provided by \pkg{nlmeODE}, parameters can be grouped in fixed effects (same parameters over all conditions) and random effects (parameters are different between conditions).

In this paper we present \pkg{dMod}, an \proglang{R} package on dynamic modeling and parameter estimation. 
The package is grown over the last years and reflects many developments and lessons learned from our research projects with time-resolved experimental data from Systems Biology. They form the basis of the development of \pkg{dMod} and are mostly referenced in the following paragraphs.\\
The aim and core functionality of \pkg{dMod} is
\begin{enumerate}[(i)]
	\item facilitated set-up of dynamic models with automated \proglang{C} code generation for fast simulation of model predictions and model sensitivities,
	\item flexible set-up of general parameter transformations (explicit or implicit) and observation functions, allowing for the implementation of multiple experimental conditions similar to mixed-effects modeling,
	\item parameter estimation based on trust-region optimization of the negative log-likelihood, making use of the sensitivity equations of the dynamic system and of symbolic derivatives of the observation- and parameter transformation functions,
	\item identifiability and uncertainty analysis based on the profile-likelihood method to determine confidence intervals for parameters and predictions.
\end{enumerate}

@@

\subsection{Simulation and prediction}
These processes involved in bile acid transport give rise to the differential equations and corresponding flowchart presented in Figure~\ref{fig:flowchart}. 
\begin{figure}[ht]
	\centering
	\begin{minipage}{0.55\textwidth}
		\textbf{A}
		\begin{equation*}
			\begin{aligned}
				\frac{\texttt d}{ \texttt{dt} } \verb!TCA_buffer! &\texttt{ = } \verb!-import * TCA_buffer!\\
				& \quad \verb! + export_sinus * TCA_cell!\\
				& \quad \verb! + reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cana! &\texttt{ = } \verb!export_cana * TCA_cell!\\
				& \quad \verb! - reflux * TCA_cana!\\
				\frac{\texttt d}{ \texttt{dt}} \verb!TCA_cell! &\texttt{ = } \verb!import * TCA_buffer!\\
				& \quad \verb! - export_sinus * TCA_cell!\\
				&\quad \verb! - export_cana * TCA_cell!.
			\end{aligned}
		\end{equation*}
	\end{minipage}
	\begin{minipage}{0.44\textwidth}
		\textbf{B}\\
	\includegraphics[width = \textwidth]{images/flowchart}
	\end{minipage}
	\caption{Differential equations and flowchart of the reaction network. Taurocholic acid,  TCA, is transported between three compartments by four different processes. (A) Assuming mass-action kinetics, the three dynamic states satisfy a set of coupled differential equations. (B) The equations are visualized in a flowchart. }
	\label{fig:flowchart}
\end{figure}
Each transportation process is modeled by mass-action kinetics. A possible implementation in \pkg{dMod} is:
\begin{CodeChunk}
	\begin{CodeInput}
  reactions <- NULL
  reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell",
  			 rate = "import*TCA_buffer",
  			 description = "Uptake")
  reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer",
  			 rate = "export_sinus*TCA_cell",
  			 description = "Sinusoidal export")
  reactions <- addReaction(reactions, "TCA_cell", "TCA_cana",
  			 rate = "export_cana*TCA_cell",
  			 description = "Canalicular export")
  reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer",
  			 rate = "reflux*TCA_cana",
  			 description = "Reflux into the buffer")
	\end{CodeInput}
\end{CodeChunk}
The reactions are translated into the ODE model object:
\begin{CodeChunk}
	\begin{CodeInput}
  mymodel <- odemodel(reactions, modelname = "bamodel")
	\end{CodeInput}
\end{CodeChunk}
Finally, the prediction function is generated:
\begin{CodeChunk}
	\begin{CodeInput}
  x <- Xs(mymodel, condition = NULL)
	\end{CodeInput}
\end{CodeChunk}
The reactions are collected in an \code{eqnlist} object. The \code{odemodel()} command composes the single reactions to an ODE system and auto-generates the \proglang{C} code which is used by the \pkg{deSolve} package to evaluate the ODE. Prediction functions are generated by the \code{Xs()} command. The usage of the prediction function is illustrated by the following code chunk. Time points are defined between 0 and 50, numeric values are assigned to all model parameters. Note that the first three parameters refer to state initial values: 
\begin{CodeChunk}
\begin{CodeInput}
  times <- seq(0, 50, .1)
  pars <- c(TCA_buffer = 1,          
            TCA_cell = 0,            
	    TCA_cana = 0,            
	    import = 0.2,
	    export_sinus = 0.2,
	    export_cana = 0.04,
	    reflux = 0.1)
\end{CodeInput}
\end{CodeChunk}
Finally the prediction function is called and both, the prediction and the sensitivities, are plotted, shown in Figure~\ref{fig:prediction}:
\begin{CodeChunk}
\begin{CodeInput}
  out <- x(times, pars)
  plot(out)

  outSens <- getDerivs(x(times, pars[4:7], fixed = pars[1:3]))
  plot(outSens)
\end{CodeInput}
\end{CodeChunk}

@@

First, observables are defined by an \code{eqnvec} object from which an observation function \code{g} is generated by the \code{Y()} command. The \code{Y()} command needs to be informed which of the symbols are variables (dynamic states) or parameters. Conveniently, \code{Y()} can retrieve this information from the prediction function \code{x} it is built on, or parse an \code{eqnvec} or \code{eqnlist} such as \code{reactions}.
%For convenience, this can be done passing the \code{reactions} that underly our ODE model.
\begin{CodeChunk}
\begin{CodeInput}
  observables <- eqnvec(
    buffer = "s*TCA_buffer",
    cellular = "s*(TCA_cana + TCA_cell)"
  )

  g <- Y(observables, f = reactions, condition = NULL,
         compile = TRUE, modelname = "obsfn")
\end{CodeInput}
\end{CodeChunk}
Observation functions link internal to observable states. Thus, providing values for the model parameters, the observation function can be used to simulate the outcome of an experiment. Adding noise to the prediction, experimental data is simulated.
In the following, we will simulate the outcome of an \textit{efflux experiment}.
The experiment starts with all TCA concentrations in steady state, such as shown in Figure~\ref{fig:prediction} after $t = 50$. To initiate the efflux, the buffer is replaced by TCA-free buffer, i.e.~\code{TCA_buffer = 0}. This translates into the following initial parameter values:
%, i.e. time point $t = 0$ reflects the distribution of TCA after uptake, as shown in Figure~\ref{fig:prediction} after $t = 50$. Hence, we reset some of the parameter values. Initial values are set to steady state values. The incubation buffer is replaced by fresh buffer, meaning that \code{TCA_buffer} is set to 0. Finally, we assume that \code{s = 1000}. The prediction of the observables is made by concatenation of \code{g} and \code{x} via the concatenation operator \code{"*"} being defined in \pkg{dMod} for prediction, observation and parameter functions.
\begin{CodeChunk}
\begin{CodeInput}
  pars["TCA_cell"] <- 0.3846154
  pars["TCA_cana"] <- 0.1538462
  pars["TCA_buffer"] <- 0
  pars["s"] <- 1e3
\end{CodeInput}
\end{CodeChunk}
The predicted dynamics of the system's internal and observable states is obtained by evaluation of the concatenated prediction function $g\circ x$, formulated as \code{g*x} in \pkg{dMod}. The scaling parameter \code{s} is set to 1000.
\begin{CodeChunk}
\begin{CodeInput}
  out <- (g*x)(times, pars, conditions = "standard")
\end{CodeInput}
\end{CodeChunk}
Since \code{g} and \code{x} have been generated as generic functions, i.e.~\code{condition = NULL}, we can assign the output to a condition of our choice, in this case ``standard''.
The predicted noiseless observation is obtained by considering the observable states only at the time point of observation \code{timesD}.
%Finally, we obtain the simulated data by taking a subset of the prediction.
\begin{CodeChunk}
\begin{CodeInput}
  timesD <- c(0.1, 1, 3, 7, 11, 15, 20, 41)
  datasheet <- subset(as.data.frame(out),
                      time %in% timesD & name %in% names(observables))
\end{CodeInput}
\end{CodeChunk}
